The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/ReviewResponses.txt

=========================================================================
Revew Responses for R Journal paper 2023-15.

We have created this text file with the content of both referees'
reports and edited our responses to comments under headings

@@

   Response X:YY

where X indicates the referee (1 or 2) and YY is an identifying
number. If X=0, the response is something we felt should be corrected.

@@

Response 0:01  We noticed a factor of 2 error in the equations
    in the discussion of the background to methods and have
    (hopefully!) corrected them.

Review 1

Referee report of “A Comparison of R Tools for Nonlinear Least
Squares Modeling”.
ID: 2023-15
General
This is a long paper that covers many topics related to nonlinear least 
squares fitting with R. This mixture of many topics can be confusing and 
I think the paper would benefit from a clearer definition of its purpose 
and focus. Once a clearer focus is established it will be easier to decide 
which topics are relevant to that purpose.

Please see Response 2:02 which responds to a similar concern.

@@

Some of the discussion in the “Jacobian code in selfStart models” section 
(p. 9) of this paper claims that the plinear fit followed by a fit using 
the default approach would be difficult to maintain and could be very slow
but that has not been the case. It is known (and, in fact, stated in the 
documentation for SSlogis) that the call to the default nls algorithm will 
always declare convergence immediately.

Response 1:03 Can we have a pointer to the documentation of the immediate 
      convergence? We have checked the documentation ?SSlogis, as well as 
      selfStart and getInitial, but do not see a mention in R 4.3.1.

      Interim action: We have reworded the paragraph starting
	"In considering `selfStart` models,"
      to reflect the comments of the referee and to provide a reference
      to the source of the starting approximations we used in SSlogisJN().

@@

Recommendations

As stated in the “General” section, I think the purpose of the paper 
should be clarified. If the purpose is to introduce the nlsr package 
and contrast it with the existing implementation of nls in the stats
package, then it might be helpful to understand why nls was implemented 
the way it was.

I hope that this unconventional referee’s report is of use to the authors 
and the editor.

Response 1:05 The historical context given in the referee's report could 
    be very useful to the R community.
    Would the editors consider asking that Referee 1 consider recasting
    the material as a companion note?

    See also 2:01 and 2:02 for clarification of purpose.

@@

Response 2:03 "port" cases added. The "appears unfinished" comment in the 
       documentation led us to avoid its use. This is the sort of issue we 
       would very much like to see resolved, since "port" is the only tool
       for bounds constraints.

Also, the model setup (Logistic3T) for which gsl_nls() fails can easily be 
made to converge correctly by changing either the algorithm or the scaling 
method, see the Illustrating code section. My advice is to provide a more 
comprehensive comparison of the available NLS solvers, or to pick a different
data example that better highlights the benefits of nlsr. For instance, trying 
to solve the BoxBOD regression problem 
(https://www.itl.nist.gov/div898/strd/nls/data/boxbod.shtml), 
nls() fails with starting values θ1 = 1, θ2 = 1 using both the Gauss-Newton 
and Port algorithm, whereas nlsr::nlxb() converges correctly.

Response 2:04 We have added a paragraph

	We consider the code inside package `nlsr` to be competently coded, 
	with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
	in common with `minpack.lm` and `glsnls`. The success rate is high,
        but `nlsr` has its premature terminations failures too. Our principal 
        message is that we obtain diagnostic information in the form of the 
        singular values of the Jacobian that allows us to try alternative starts, 
        algorithms or programs in cases where there are indications of early termination.

@@

    We have added two small comments on recommended usage of bounds. One
    is to control for wild excursions in the optimization trajectory, as 
    we believe this is sensible when bounds constraints are available, 
    as in nlsr::nlxb. We also note that in cases where a simple and 
    obvious transformation is available, such as squaring a parameter to 
    ensure positivity, this is likely a better approach than imposing bounds.
    However, when there are more general bounds, transformations become more
    tricky to apply, and may de-scale a problem. We have considered the
    transfinite function used in CRAN package `adagio`, but consider that
    this needs more extensive investigation before recommending in general.

5. The section Programming language makes the argument that keeping to a 
single programming language can allow for easier maintenance and upgrades, 
and that the performance penalty for using code entirely in R has been 
much reduced in recent years. I agree that there is not much of a 
performance penalty when solving small problems, such as the data examples 
included in the article. However, when solving problems using a large amount 
of data or a large number of parameters, I have experienced that the run-time 
(as well as memory usage) can become an actual bottleneck. As an example, 
consider the Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) 
included in the Illustrating code section.

Response 2:08  In that we are in general agreement with this comment, we have added
    some clarification to advise that it is worthwhile being aware of how long
    calculations are taking.

    The referee's comments and Illustrating Code led us to try the Pen1 problem
    (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
    to which one of us recently added Hessian code). This showed that the R code
    of nlsr::nlfb() was generally slower by a small factor than the compiled 
    Fortran of minpack.lm::nls.lm. However, the all-R code optimx::Rcgmin was 
    much faster for larger problems -- it does not build a matrix!

### 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/StateOfRToolsForNLS.Rmd
Prep for resubmission

---
title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-3-2"
output: 
  pdf_document:
    keep_tex: yes
    toc: no
  word_document:
    toc: no
bibliography: ./StateOfRToolsForNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Our Google Summer of Code project "Improvements to `nls()`" considered the features 
and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, in particular `nlsr` (@nlsr2023manual), provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

Besides the base-R `nls()` function, we consider some packages that 
are available on the CRAN repository. Of these, we will pay particular 
attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers. We will 
not pursue tools in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub and Gitlab.

@@

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require an 
examination of the code. `nlsr` terminates when the number of residual or 
Jacobian evaluations (`feval` and `jeval`) exceed set limits, and we prefer larger 
limits (`jemax = 5000` is the default) than the `maxiter = 50` of `nls()` to avoid 
stopping early. This may cause overly long run-times for `nlsr`.

@@

Bounds can also be helpful to avoid unanticipated excursions of the minimization
trajectory into inadmissible parameter regions. Such bounds should not be too
tight, as this may hinder progress to an optimum by causing the methods to 
bump along the constraint in many cycles. Our aim is to avoid rubbish, not
specify where the solution will be, and we hope the solution is complicated
by having a parameter on a bound.

@@

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls`, and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods which we
have not investigated.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants. In general, the Levenberg-Marquardt stabilization 
is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors. Successful
incorporation of Marquardt stabilization into `nls()` would be a major improvement.

@@

There are problems where use of R will be slower. For example, a
"large" case of the Penalty 1 test problem of @More1981TUO will take
noticeable time to solve. Problems in large numbers of parameters
create large computational matrices for the Jacobian or its inner product,
which Gauss-Newton methods create and process. 
This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
the Penalty 1 problem of various sizes, where the latter, which uses 
compiled Fortran internally, was moderately faster. However, the method
"Rcgmin" from package `optimx` (@p-optimx) proved to be much 
faster than either, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods, and the pattern of
such timings varies much less regularly with problem size for CG methods.

@@

Some models have several common parameters and 
others that are tied to particular cases. 
The **man** file for `nls()` includes an example of a situation in which
parameters are indexed but which uses the "plinear" option as an added complication.
Running this example reveals that the answers for the parameters are not indexed
as in a vector. 
<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
<!-- This is because the model -->
<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
This is no doubt because programming for indexed parameters is challenging.
<!-- However, we believe the current structure could cause confusion and error, and  -->
<!-- propose an alternative approach below. -->
We note that there are capabilities in packages for mixed-effects modeling
such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (p-lme4) for 
estimating models of such type.

@@

Such goals align with the aims of **unit testing** 
(@HWtestthat11, @HWdevtools21, the conventional R package testing tools, and 
a number of blog posts found by searching "unit testing in R").
In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.

>>>>

---
title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-7-16"
output: 
  pdf_document:
    keep_tex: yes
    toc: no
  word_document:
    toc: no
bibliography: ./StateOfRToolsForNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Our Google Summer of Code project "Improvements to `nls()`" considered the features 
and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` offers.
Various packages, in particular `nlsr` (@nlsr2023manual), provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

Besides the base-R `nls()` function, we consider some packages that 
are available in the CRAN repository. Of these, we will pay particular 
attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers. We will 
not pursue tools in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub and Gitlab.

@@

Termination after what may be considered excessive computation is also important.
`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require an 
examination of the code. `nlsr` terminates when the number of residual or 
Jacobian evaluations (`feval` and `jeval`) exceed set limits, and we prefer larger 
limits (`jemax = 5000` is the default) than the `maxiter = 50` of `nls()` to avoid 
stopping early. This may cause overly long run-times for `nlsr`, but our work often
involves problems that are sufficiently difficult that we want to be sure we do not
terminate early.

@@

Bounds can also be helpful to avoid unanticipated excursions of the minimization
trajectory into inadmissible parameter regions. Such bounds should not be too
tight, as this may hinder progress to an optimum by causing the methods to 
bump along the constraint in many cycles. Our aim is to avoid rubbish, not
specify where the solution will be, and we hope the solution is not complicated
by having a parameter on a bound.

@@

All four major tools illustrated solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls`, and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods which we
have not investigated beyond one example.
Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
algorithm variants, principally for the benefit of program developers. 
In general, the Levenberg-Marquardt stabilization 
is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors. A tidy
incorporation of Marquardt stabilization into `nls()` would be a major improvement.

@@

There are, of course, problems where use of R code will be slower. For example, a
"large" case of the Penalty 1 test problem of @More1981TUO will take
noticeable time to solve. Problems in large numbers of parameters
create large computational matrices for the Jacobian or its inner product,
which Gauss-Newton methods create and process. 
This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
the Penalty 1 problem of various sizes, where the latter, which uses 
compiled Fortran internally, was moderately faster. However, the method
"Rcgmin" from package `optimx` (@p-optimx) proved to be much 
faster than either, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods, and the pattern of
such timings varies much less regularly with problem size for CG methods.

@@

Some models have several common parameters and 
others that are tied to particular cases. 
The **man** file for `nls()` includes an example of a situation in which
parameters are indexed but which uses the "plinear" option as an added complication.
Running this example reveals that the answers for the parameters are not indexed
as in a vector. 
<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
<!-- This is because the model -->
<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
This is no doubt because programming for indexed parameters is challenging.
<!-- However, we believe the current structure could cause confusion and error, and  -->
<!-- propose an alternative approach below. -->
We note that there are capabilities in packages for mixed-effects modeling
such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (@p-lme4) for 
estimating models of such type.

@@

Such goals align with the aims of **unit testing** 
(@HWtestthat11, @HWdevtools21, the conventional R package testing tools, and 
a number of blog posts found by searching "unit testing in R").
In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare*.
A primary design objective of this is to allow the summarisation of multiple tests in
a compact output. The prototype has a vignette to illustrate its use.
